---
title: "Bootstrapped Bounding Analysis"
format: html
---

This document isolates the computationally intensive bootstrapping validation for the Fosse & Winship Bounded APC Analysis, following Rohrer (2025).

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(boot)

library(haven)
df_raw <- read_dta("sppa1982-2012_stata.dta", col_select = c("year", "age", "gender", "education", "race", "jazz", "classical", "opera", "musical", "ballet", "artmuseum", "craft_fair", "park", "weight_normalized"))

outcomes <- c("jazz", "classical", "opera", "musical", "ballet", "artmuseum", "craft_fair", "park")

get_cohort_bins <- function(cohort_val, width, start_yr = 1885, end_yr = 1995) {
  breaks <- seq(start_yr, end_yr, by = width)
  bins <- cut(cohort_val, breaks = breaks, right = FALSE, labels = FALSE)
  bin_labels <- breaks[bins]
  if (width == 5) {
    bin_labels[bin_labels == 1885] <- 1890
  }
  return(bin_labels)
}

df_clean <- df_raw |>
  mutate(across(all_of(outcomes), ~ case_when(
    . == 1 ~ 1,
    . == 2 ~ 0,
    TRUE ~ NA_real_
  ))) |>
  filter(!is.na(craft_fair)) |>
  mutate(
    woman = if_else(gender == 2, 1, 0),
    college = if_else(education > 4, 1, 0),
    white = if_else(race == 1, 1, 0),
    black = if_else(race == 2, 1, 0),
    agecat = cut(age, breaks = seq(15, 100, by = 5), right = FALSE, labels = FALSE),
    cohort = year - age,
    cohort5 = get_cohort_bins(cohort, 5),
    omnivore_score = rowSums(across(all_of(outcomes)), na.rm = TRUE)
  ) |>
  filter(age >= 15 & age <= 95)

# Get the initial df_bounds_omni 
fit_apc_omni <- glm(
  omnivore_score ~ as.factor(agecat) + as.factor(year) + as.factor(cohort5) + woman + college + white + black,
  family = poisson(link = "log"), 
  data = df_clean
)
coefs_omni <- coef(fit_apc_omni)

cohort_years <- seq(1890, 1990, by = 5)
cohort_coefs_omni <- c(0, coefs_omni[grep("as.factor\\(cohort5\\)", names(coefs_omni))])
names(cohort_coefs_omni) <- cohort_years

age_coefs_omni <- c(0, coefs_omni[grep("as.factor\\(agecat\\)", names(coefs_omni))])
names(age_coefs_omni) <- seq(15, 95, by = 5)

period_coefs_omni <- c(0, coefs_omni[grep("as.factor\\(year\\)", names(coefs_omni))])
names(period_coefs_omni) <- c(1982, 1985, 1992, 2002, 2008, 2012)

s_upper_age_o <- as.numeric((age_coefs_omni["50"] - age_coefs_omni["80"]) / 30)
s_upper_period_o <- as.numeric((period_coefs_omni["2012"] - period_coefs_omni["1982"] + 1.5) / 30)
s_lower_period_o <- as.numeric((period_coefs_omni["2012"] - period_coefs_omni["1982"]) / 30)

s_max_o <- min(s_upper_age_o, s_upper_period_o)
s_min_o <- s_lower_period_o

df_bounds_omni <- tibble(
  cohort = cohort_years,
  effect_min = cohort_coefs_omni + s_min_o * (cohort - (1982 - 15)),
  effect_zero = cohort_coefs_omni,
  effect_max = cohort_coefs_omni + s_max_o * (cohort - (1982 - 15))
)
```

### Incorporating Sampling Variability via Bootstrapping

As noted by Rohrer (2025), the bounding procedure incorporates uncertainty regarding the age-period-cohort identification problem but does not inherently account for statistical sampling error. To generate confidence intervals around our bounded cohort effects and test whether the estimated bounds significantly differ from zero, we can bootstrap the entire bounding procedure.

```{r bootstrap-bounds, cache=TRUE, message=FALSE, warning=FALSE}
library(boot)

# Define the bootstrapping function
boot_bounds <- function(data, indices) {
  d <- data[indices, ]
  fit <- glm(
    omnivore_score ~ as.factor(agecat) + as.factor(year) + as.factor(cohort5) + woman + college + white + black,
    family = poisson(link = "log"), 
    data = d
  )
  coefs <- coef(fit)
  
  # Cohort vector (extract existing, set missing to NA or 0)
  c_years <- seq(1890, 1990, by = 5)
  c_coefs <- rep(NA, length(c_years))
  names(c_coefs) <- c_years
  c_coefs["1890"] <- 0 # reference
  for (cy in c_years[-1]) {
    nm <- paste0("as.factor(cohort5)", cy)
    if (nm %in% names(coefs)) c_coefs[as.character(cy)] <- coefs[nm]
  }
  
  # Age vector
  a_years <- seq(15, 95, by = 5)
  a_cats <- 1:17
  a_coefs <- rep(NA, length(a_years))
  names(a_coefs) <- a_years
  a_coefs["15"] <- 0 # reference
  for (i in 2:17) {
    nm <- paste0("as.factor(agecat)", i)
    if (nm %in% names(coefs)) a_coefs[as.character(a_years[i])] <- coefs[nm]
  }
  
  # Period vector
  p_years <- c(1982, 1985, 1992, 2002, 2008, 2012)
  p_coefs <- rep(NA, length(p_years))
  names(p_coefs) <- p_years
  p_coefs["1982"] <- 0 # reference
  for (py in p_years[-1]) {
    nm <- paste0("as.factor(year)", py)
    if (nm %in% names(coefs)) p_coefs[as.character(py)] <- coefs[nm]
  }
  
  A_ref <- 15
  P_ref <- 1982
  C_ref <- P_ref - A_ref
  
  s_upper_age <- as.numeric((a_coefs["50"] - a_coefs["80"]) / 30)
  s_upper_period <- as.numeric((p_coefs["2012"] - p_coefs["1982"] + 1.5) / 30)
  s_lower_period <- as.numeric((p_coefs["2012"] - p_coefs["1982"]) / 30)

  
  s_max <- min(s_upper_age, s_upper_period, na.rm=TRUE)
  s_min <- s_lower_period
  
  eff_min <- c_coefs + s_min * (c_years - C_ref)
  eff_max <- c_coefs + s_max * (c_years - C_ref)
  
  return(c(eff_min, eff_max))
}

# Run bootstrap (R=100 for computational efficiency, can be increased for final manuscript)
set.seed(42)
boot_results <- boot(data = df_clean, statistic = boot_bounds, R = 100)

# Extract 95% CIs
c_years <- seq(1890, 1990, by = 5)
n_c <- length(c_years)

ci_min_lower <- apply(boot_results$t[, 1:n_c], 2, quantile, probs = 0.025, na.rm=TRUE)
ci_min_upper <- apply(boot_results$t[, 1:n_c], 2, quantile, probs = 0.975, na.rm=TRUE)

ci_max_lower <- apply(boot_results$t[, (n_c+1):(2*n_c)], 2, quantile, probs = 0.025, na.rm=TRUE)
ci_max_upper <- apply(boot_results$t[, (n_c+1):(2*n_c)], 2, quantile, probs = 0.975, na.rm=TRUE)

df_bounds_boot <- df_bounds_omni |>
  mutate(
    min_ci_low = ci_min_lower,
    min_ci_high = ci_min_upper,
    max_ci_low = ci_max_lower,
    max_ci_high = ci_max_upper
  )

p_bounds_boot <- ggplot(df_bounds_boot, aes(x = cohort)) +
  geom_ribbon(aes(ymin = min_ci_low, ymax = min_ci_high), fill = "blue", alpha = 0.15) +
  geom_ribbon(aes(ymin = max_ci_low, ymax = max_ci_high), fill = "red", alpha = 0.15) +
  geom_line(aes(y = effect_min, color = "Lower Bound"), linewidth = 1, linetype = "dashed") +
  geom_line(aes(y = effect_max, color = "Upper Bound"), linewidth = 1, linetype = "dotdash") +
  geom_line(aes(y = effect_zero, color = "Baseline Reference"), linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "solid", alpha = 0.3) +
  scale_color_manual(values = c("Lower Bound" = "blue", "Upper Bound" = "red", "Baseline Reference" = "black")) +
  theme_minimal() +
  labs(
    title = "Bootstrapped Bounded Cohort Effects",
    subtitle = "95% CIs for Fosse & Winship Bounding Approach",
    x = "Birth Cohort",
    y = "Cohort Effect (Log-Count)",
    color = "Boundary"
  ) +
  theme(legend.position = "bottom")

p_bounds_boot

ggsave("Plots/cohort_bounds_bootstrapped.png", p_bounds_boot, width = 8, height = 5, dpi = 300)
```
