---
title: "Rohrer Bootstrapping: Bounded Cohort Effects"
format: html
---

## Setup
```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(haven)
library(boot)

# Load data (assuming df_clean is already saved or we can recreate it)
# For this script we will recreate df_clean to be standalone
df_raw <- read_dta("sppa1982-2012_stata.dta", col_select = c(
  "persid", "year", "age", "gender", "education", "race", "hispanic",
  "jazz", "classical", "opera", "musical", "ballet", "artmuseum", "craft_fair", "park",
  "weight_normalized"
))

outcomes <- c("jazz", "classical", "opera", "musical", "ballet", "artmuseum", "craft_fair", "park")

df_clean <- df_raw |>
  mutate(across(all_of(outcomes), ~ case_when(. == 1 ~ 1, . == 2 ~ 0, TRUE ~ NA_real_))) |>
  filter(!is.na(craft_fair)) |>
  mutate(
    woman = if_else(gender == 2, 1, 0),
    college = if_else(education > 4, 1, 0),
    white = if_else(race == 1, 1, 0),
    black = if_else(race == 2, 1, 0),
    agecat = cut(age, breaks = seq(15, 100, by = 5), right = FALSE, labels = FALSE),
    cohort = year - age
  ) |>
  filter(age >= 15 & age <= 95)

get_cohort_bins <- function(cohort_val, width, start_yr = 1885, end_yr = 1995) {
  breaks <- seq(start_yr, end_yr, by = width)
  bins <- cut(cohort_val, breaks = breaks, right = FALSE, labels = FALSE)
  bin_labels <- breaks[bins]
  if (width == 5) bin_labels[bin_labels == 1885] <- 1890
  return(bin_labels)
}

df_clean <- df_clean |>
  mutate(
    cohort5 = get_cohort_bins(cohort, 5),
    omnivore_score = rowSums(across(all_of(outcomes)), na.rm = TRUE)
  )

# Precompute df_bounds_omni for plotting later
c_years <- seq(1890, 1990, by = 5)
df_bounds_omni <- tibble(cohort = c_years)
```

### Incorporating Sampling Variability via Bootstrapping

As noted by Rohrer (2025), the bounding procedure incorporates uncertainty regarding the age-period-cohort identification problem but does not inherently account for statistical sampling error. To generate confidence intervals around our bounded cohort effects and test whether the estimated bounds significantly differ from zero, we can bootstrap the entire bounding procedure.

```{r bootstrap-bounds, cache=TRUE, message=FALSE, warning=FALSE}
# Define the bootstrapping function
boot_bounds <- function(data, indices) {
  d <- data[indices, ]
  fit <- glm(
    omnivore_score ~ as.factor(agecat) + as.factor(year) + as.factor(cohort5) + woman + college + white + black,
    family = poisson(link = "log"), 
    data = d
  )
  coefs <- coef(fit)
  
  c_years <- seq(1890, 1990, by = 5)
  c_coefs <- rep(NA, length(c_years))
  names(c_coefs) <- c_years
  c_coefs["1890"] <- 0 # reference
  for (cy in c_years[-1]) {
    nm <- paste0("as.factor(cohort5)", cy)
    if (nm %in% names(coefs)) c_coefs[as.character(cy)] <- coefs[nm]
  }
  
  a_years <- seq(15, 95, by = 5)
  a_cats <- 1:17
  a_coefs <- rep(NA, length(a_years))
  names(a_coefs) <- a_years
  a_coefs["15"] <- 0 # reference
  for (i in 2:17) {
    nm <- paste0("as.factor(agecat)", i)
    if (nm %in% names(coefs)) a_coefs[as.character(a_years[i])] <- coefs[nm]
  }
  
  p_years <- c(1982, 1985, 1992, 2002, 2008, 2012)
  p_coefs <- rep(NA, length(p_years))
  names(p_coefs) <- p_years
  p_coefs["1982"] <- 0 # reference
  for (py in p_years[-1]) {
    nm <- paste0("as.factor(year)", py)
    if (nm %in% names(coefs)) p_coefs[as.character(py)] <- coefs[nm]
  }
  
  A_ref <- 15
  P_ref <- 1982
  C_ref <- P_ref - A_ref
  
  s_upper_age <- as.numeric((a_coefs["50"] - a_coefs["80"]) / 30)
  s_upper_period <- as.numeric((p_coefs["2012"] - p_coefs["1982"] + 1.5) / 30)
  s_lower_period <- as.numeric((p_coefs["2012"] - p_coefs["1982"]) / 30)

  s_max <- min(s_upper_age, s_upper_period, na.rm=TRUE)
  s_min <- s_lower_period
  
  eff_min <- c_coefs + s_min * (c_years - C_ref)
  eff_max <- c_coefs + s_max * (c_years - C_ref)
  
  return(c(eff_min, eff_max, c_coefs, s_min, s_max))
}

# Run bootstrap (R=100 for computational efficiency)
set.seed(42)
boot_results <- boot(data = df_clean, statistic = boot_bounds, R = 100)

# Extract 95% CIs
n_c <- length(c_years)

ci_min_lower <- apply(boot_results$t[, 1:n_c], 2, quantile, probs = 0.025, na.rm=TRUE)
ci_min_upper <- apply(boot_results$t[, 1:n_c], 2, quantile, probs = 0.975, na.rm=TRUE)

ci_max_lower <- apply(boot_results$t[, (n_c+1):(2*n_c)], 2, quantile, probs = 0.025, na.rm=TRUE)
ci_max_upper <- apply(boot_results$t[, (n_c+1):(2*n_c)], 2, quantile, probs = 0.975, na.rm=TRUE)

effect_zero_mean <- apply(boot_results$t[, (2*n_c+1):(3*n_c)], 2, mean, na.rm=TRUE)
effect_min_mean <- apply(boot_results$t[, 1:n_c], 2, mean, na.rm=TRUE)
effect_max_mean <- apply(boot_results$t[, (n_c+1):(2*n_c)], 2, mean, na.rm=TRUE)

df_bounds_boot <- df_bounds_omni |>
  mutate(
    effect_zero = effect_zero_mean,
    effect_min = effect_min_mean,
    effect_max = effect_max_mean,
    min_ci_low = ci_min_lower,
    min_ci_high = ci_min_upper,
    max_ci_low = ci_max_lower,
    max_ci_high = ci_max_upper
  )

p_bounds_boot <- ggplot(df_bounds_boot, aes(x = cohort)) +
  geom_ribbon(aes(ymin = min_ci_low, ymax = min_ci_high), fill = "blue", alpha = 0.15) +
  geom_ribbon(aes(ymin = max_ci_low, ymax = max_ci_high), fill = "red", alpha = 0.15) +
  geom_line(aes(y = effect_min, color = "Lower Bound"), linewidth = 1, linetype = "dashed") +
  geom_line(aes(y = effect_max, color = "Upper Bound"), linewidth = 1, linetype = "dotdash") +
  geom_line(aes(y = effect_zero, color = "Baseline Reference"), linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "solid", alpha = 0.3) +
  scale_color_manual(values = c("Lower Bound" = "blue", "Upper Bound" = "red", "Baseline Reference" = "black")) +
  theme_minimal() +
  labs(
    title = "Bootstrapped Bounded Cohort Effects",
    subtitle = "95% CIs for Fosse & Winship Bounding Approach",
    x = "Birth Cohort",
    y = "Cohort Effect (Log-Count)",
    color = "Boundary"
  ) +
  theme(legend.position = "bottom")

p_bounds_boot

ggsave("Plots/cohort_bounds_bootstrapped.png", p_bounds_boot, width = 8, height = 5, dpi = 300)
```
